A Standardized Effect Size for Evaluating the Strength of Phylogenetic Signal, and Why Lambda is not Appropriate

### 2020-04-28 Manuscript/CoverPage-Evolution.Rmd
vsn 3

---
title: MISUSES AND INACCURACY PATTERNS OF PAGEL'S LAMBDA ML ESTIMATION

@@

\hfill\break
**Keywords**: Pagel's lambda, phylogenetic signal
\hfill\break

**Short Title**: Problems with Pagel's lambda \hfill\break

>>>>

---
title: A Standardized Effect Size for Evaluating the Strength of Phylogenetic Signal, and Why Lambda is not Appropriate

@@

\hfill\break

**Keywords**: phylogenetic signal, effect size, Pagel's lambda \hfill\break

**Short Title**: An Effect Size for Phylogenetic Signal \hfill\break

### 2020-04-28 Manuscript/Manuscript-3.Rmd
vsn 3

**Keywords**: phylogenetic signal, effect size, Pagel's lambda \hfill\break

**Short Title**: An Effect Size for Phylogenetic Signal \hfill\break

@@

{conclusion holds: interpreting the regression is not appreciably different (in terms of slopes and f values)}

@@

Evolutionary biologists commonly seek to describe the relative strength of phylogenetic signal in phenotypic traits, to determine the extent to which shared evolutionary history has influenced trait covariation among taxa. This is often accomplished by interpreting empirical estimates of $\lambda$; with smaller values signifying 'weak' phylogenetic signal, while larger values are interpreted as 'strong' phylogenetic signal [e.g., @DeMeester2019; @Pintanel2019;@Su2019]. Other approaches for interpreting $\lambda$ are more statistical. For instance, some have evaluated whether the observed $\lambda$ differs from some expected value through the use of confidence intervals [@Vandelook2019] or by performing likelihood ratio tests that compare the observed model fit to that obtained when $\lambda=0$ or $\lambda=1$ [@Freckleton_et_al2002; @Cooper2010; @Bose2019]. Additionally, qualitative comparisons of $\lambda$ estimates obtained from multiple phenotypic traits have been used to infer whether the strength of phylogenetic signal is greater in one trait as compared to another [e.g., @Liu2019; @Bai2019]. Indeed, statements regarding the strength of phylogenetic signal based on $\lambda$ are rather common in the evolutionary literature. We conducted a literature survey in Google.scholar and found that of the 204 papers published in 2019 that estimated and reported Pagel's $\lambda$, 40% interpreted the strength of phylogenetic signal for at least one phenotypic trait. Additionally, because nearly half of the 1,572 $\lambda$ values reported were near the limits of the parameter (Figure 1), this percentage is even higher, as biological interpretation of phylogenetic signal at the limits of $\lambda$ are known. \hfill\break

@@

In this study, we evaluate the precision of Pagel's $\lambda$ in estimating known levels of phylogenetic signal in phenotypic data. We use computer simulations with differing numbers of species, differently shaped phylogenies, and differing input levels of phylogenetic signal, to explore the degree to which $\lambda$ correctly identifies known levels of phylogenetic signal, and under what circumstances. We find that while PGLS parameters (e.g., $\beta$) are accurately estimated with the inclusion of phylogenetic signal, estimates of $\lambda$ are not. We also find that estimates of $\lambda$ vary widely for a given input value of phylogenetic signal, and that the precision in estimating $\lambda$ is not constant across the range of input signal, with decreased precision when phylogenetic signal is of intermediate strength. Additionally, the same estimated values of $\lambda$ may be obtained from datasets containing vastly different input levels of phylogenetic signal. Thus, $\lambda$ is not a reliable estimate of the strength of phylogenetic signal in phenotypic data. We subsequently derive a standardized effect size for measuring the strength of phylogenetic signal in phenotypic datasets, and apply the concept to two common measures of phylogenetic signal: $\lambda$ and $Kappa$. Through simulations across a wide range of conditions, we find that the precision of effect sizes based on $\lambda$ ($Z_{\lambda}$) are less reliable than that those based on $Kappa$ ($Z_K$), implying that $Z_K$ is a more robust effect size measure. Additionally, we propose a two-sample test statistic that may be used to compare the strength of phylogenetic signal among datasets, and provide an empirical example to demonstrate its use. We conclude that estimates of phylogenetic signal using Pagel's $\lambda$ are often inaccurate, and thus interpreting strength of phylogenetic signal in phenotypic datasets based on this measure is compromised. By contrast, effect sizes obtained from $Kappa$ hold promise for characterizing phylogenetic signal, and for comparing the strength of phylogenetic signal across datasets.

@@

All analyses were performed in R v3.6.0 [@RCT] using the packages `geiger` [@Harmon2008], `caper` [@Orme2013], `phytools` [@Revell2012], and `geomorph` [@AdamsOtarola2013; @AdamsGeomorph]. R-scripts are found in the Supporting Information. \hfill\break

*Results.* We found that the precision of $\lambda_{est}$ varied widely across simulation conditions. Predictably, precision improved as the number of species increased (Figure 2). This confirmed earlier findings of Boettiger et al. [-@Boettiger_et_al2012], and adhered to parametric statistical theory. However, in many cases the set of $\lambda_{est}$ spanned nearly the entire range of possible values (e.g., $n=32$; $\lambda_{in}=0.5$: $\lambda_{est}= 0.0\to 0.985$), revealing that estimates of $\lambda$ were not a reliable indicator of input phylogenetic signal. Importantly, the precision of $\lambda_{est}$ was not uniform across all levels of phylogenetic signal, with the worst precision at intermediate levels of signal ($\lambda_{in}\approx0.5$), and improved precision as input levels approached the extremes of its range (i.e., $\lambda_{in}\rightarrow0$ & $\lambda_{in}\rightarrow1$). Thus, estimates of $\lambda$ were least reflective of the true input signal at intermediate values. Additionally, even at large levels of species richness, we found that the range of $\lambda_{est}$ still encompassed a substantial portion of possible values (e.g., $n=512$; $\lambda_{in}=0.5$: $\lambda_{est} = 0.32\to 0.68$). Likewise, the same $\lambda_{est}$ could be obtained from datasets containing vastly different input levels of phylogenetic signal (e.g., $n=512$; $\lambda_{est} = 0.5$; $\lambda_{in} = 0.25 \to 0.65$). Note that these findings were particularly unsettling when considered in light of our literature survey. Over one quarter of the $\lambda$ estimates obtained in empirical studies (421 of 1,572) were between $\lambda=0.25$ and $\lambda=0.75$ (Figure 1). Of those, 30% were interpreted directly in terms of the strength of phylogenetic signal that they represented (e.g., small, intermediate, strong signal). \hfill\break

Finally, when $\lambda$ was co-estimated with regression parameters in PGLS regression, the results of our simulations were quite similar. Here, regression parameters ($\beta$) were accurately estimated, confirming earlier findings of Revell [-@Revell2010] (see Supporting Information). Estimates of phylogenetic signal were less precise, and the spread of $\lambda_{est}$ was similar to that observed when $\lambda$ was estimated for only the dependent variable (Figure 3). Taken together, these findings reveal that $\lambda_{est}$ does not precisely characterize observed levels of phylogenetic signal in phenotypic datasets, and that biological interpretations of the strength of phylogenetic signal based on $\lambda$ may be highly inaccurate. \hfill\break

@@

The results above demonstate that $\lambda$ is not a reliable estimate of the phylogenetic signal in phenotypic data. As such, biological interpretations of the strength of phylogenetic signal, and comparisons of the magnitude of such effects across datasets, are severely compromised when based on this parameter. As an alternative, we propose that summary estimates of phylogenetic signal be converted to effect sizes for interpretation and comparison. Statistically, a standardized effect size may be found as:

@@

where $\theta_{obs}$ is the observed test statistic, $E(\theta)$ is its expected value under the null hypothesis, and $\sigma_\theta$ is its standard error [@Glass1976; @Cohen1988;@Rosenthal1994]. $Z_{\theta}$ expresses the magnitude of the effect in $\theta_{obs}$ by transforming the original test statistic to a standard normal deviate [@Glass1976; @Kelley2012]. Typically, $\theta_{obs}$ and $\sigma_\theta$ are estimated from the data, while $E(\theta)$ is obtained from the distribution of $\theta$ derived from parametric theory. However, recent advances in resampling theory [@Collyer_et_al2015a;@AdamsCollyer2016; @AdamsCollyer2019b] have shown that $E(\theta)$ and $\sigma_\theta$ may also be obtained from an empirical sampling distribution of $\theta$ obtained from permutation procedures. \hfill\break

Adams and Collyer [-@AdamsCollyer2019] recently suggested that the strength of phylogenetic signal could be represented as an effect size, based on the $Kappa$ statistic and its empirical sampling distribution from permutation. Here we formalize that suggestion, resulting in an effect size of:

@@

In this case, $\lambda_{obs}$ and $\hat\sigma_\lambda$ are empirically derived using maximum likelihood. Note that under the null hypothesis no phylogenetic signal is expected [@Freckleton_et_al2002], and thus $E(\lambda)=0$ under this condition. \hfill\break

To evaluate the utility of $Z_K$ and $Z_\lambda$ we calculated both effect sizes for the simulated datasets generated above, and summarized the precision of each using its variance ($\sigma^2_{Z_K}$ and $\sigma^2_{Z_\lambda}$). Results are found in Figure 4. Here two things are evident. First, estimates of $Z_K$ track the input phylogenetic signal in a more linear fashion than do estimates of $Z_\lambda$ (Figure 4A,B). Thus, actual changes in the strength of phylogenetic signal are reflected more evenly in the corresponding values of the effect size $Z_K$. Second, the precision of $Z_K$ is considerably more stable as compared with $Z_\lambda$. This may be seen by calculating the coefficients of variation for the set of precision estimates (i.e., $\sigma^2_{Z_K}$ and $\sigma^2_{Z_\lambda}$) across input levels of phylogenetic signal.  Here coefficients of variation in the precision of $Z_K$ were an order of magnitude smaller for than for $Z_\lambda$ (Figure 4C,D). This implied that estimates of the strength of phylogenetic signal were more reliable and robust when using $Z_K$ as compared with $Z_\lambda$. \hfill\break

@@

One important advantage of the approach advocated here is that the resulting effect sizes ($Z_K$) are dimensionless, as the units of measurement cancel out during their calculation [**S&R??**]. As a consequence, the effect sizes represent the strength of phylogenetic signal on a common and comparable scale, measured in standard deviation units, regardless of the initial units and scale of the original phenotypic variables under investigation.

###
Conceptually, this is akin to the conversion of data to their standard normal deviates prior to their use in multivariate analysis, when the original variables are expressed in different units or scales [see ]. 
###

####

mathematical transformation or standardization of the trait dimensions, so
the variables are expressed in similar units
n the variables in a multivariate data set are on different scales,

Everitt, Brian; Hothorn, Torsten J (2011), An Introduction to Applied Multivariate Analysis with R, Springer,

Also, Legendre... Adams and Collyer. 
####

Because of this standardization, the strength of phylogenetic signal across datasets has meaning

something about comparisons having meaning...

As such, characterizing the strength of phylogenetic signal with $Z_K$ facilitates a formal statistical comparison of values across continuous traits of different units, scale, or even dimensionality (e.g., univariate traits using $Kappa$, and multivariate traits using $K_{mult}$).

@@

>>>>

@@

@@

Evolutionary biologists commonly seek to describe the relative strength of phylogenetic signal in phenotypic traits, to determine the extent to which shared evolutionary history has influenced trait covariation among taxa. This is often accomplished by interpreting empirical estimates of $\lambda$; with smaller values signifying 'weak' phylogenetic signal, while larger values are interpreted as 'strong' phylogenetic signal [e.g., @DeMeester2019; @Pintanel2019;@Su2019]. Other approaches for interpreting $\lambda$ are more statistical. For instance, some have evaluated whether the observed $\lambda$ differs from some expected value through the use of confidence intervals [@Vandelook2019] or by performing likelihood ratio tests that compare the observed model fit to that obtained when $\lambda=0$ or $\lambda=1$ [@Freckleton_et_al2002; @Cooper2010; @Bose2019]. Additionally, qualitative comparisons of $\lambda$ estimates obtained from multiple phenotypic traits have been used to infer whether the strength of phylogenetic signal is greater in one trait as compared to another [e.g., @Liu2019; @Bai2019]. Indeed, statements regarding the strength of phylogenetic signal based on $\lambda$ are rather common in the evolutionary literature. For instance, of the 204 papers published in 2019 that estimated and reported Pagel's $\lambda$ (found in Google.scholar), 40% interpreted the strength of phylogenetic signal for at least one phenotypic trait. Additionally, because nearly half of the 1,572 $\lambda$ values reported were near the limits of the parameter (Figure 1), this percentage is even higher, as biological interpretation of phylogenetic signal at the limits of $\lambda$ are known. \hfill\break

@@

In this study, we evaluate the precision of Pagel's $\lambda$ for estimating known levels of phylogenetic signal in phenotypic data. We use computer simulations with differing numbers of species, differently shaped phylogenies, and differing input levels of phylogenetic signal, to explore the degree to which $\lambda$ correctly identifies known levels of phylogenetic signal, and under what circumstances. We find that estimates of $\lambda$ vary widely for a given input value of phylogenetic signal, and that the precision in estimating $\lambda$ is not constant across its range. Rather, there is decreased precision when input levels of phylogenetic signal are of intermediate strength. Additionally, the same estimated values of $\lambda$ may be obtained from datasets containing vastly different input levels of phylogenetic signal. Thus, $\lambda$ is not a reliable estimate of the strength of phylogenetic signal in phenotypic data. We then describe a standardized effect size for measuring the strength of phylogenetic signal in phenotypic datasets, and apply the concept to two common measures of phylogenetic signal: $\lambda$ and $Kappa$. Through simulations across a wide range of conditions, we find that the precision of effect sizes based on $\lambda$ ($Z_{\lambda}$) are less reliable than that those based on $Kappa$ ($Z_K$), implying that $Z_K$ is a more robust effect size measure. We also propose a two-sample test statistic that may be used to compare the strength of phylogenetic signal among datasets, and provide an empirical example to demonstrate its use. We conclude that estimates of phylogenetic signal using Pagel's $\lambda$ are often inaccurate, and thus interpreting strength of phylogenetic signal in phenotypic datasets based on this measure is compromised. By contrast, effect sizes obtained from $Kappa$ hold promise for characterizing phylogenetic signal, and for comparing the strength of phylogenetic signal across datasets.

@@

All analyses were performed in R v3.6.0 [@RCT] using the packages `geiger` [@Harmon2008], `caper` [@Orme2013], `phytools` [@Revell2012], and `geomorph` 3.2.1 [@AdamsOtarola2013; @AdamsGeomorph]. R-scripts are found in the Supporting Information. \hfill\break

*Results.* We found that the precision of $\lambda_{est}$ varied widely across simulation conditions. Predictably, precision improved as the number of species increased (Figure 2). This confirmed earlier findings of Boettiger et al. [-@Boettiger_et_al2012], and adhered to parametric statistical theory. However, in many cases the set of $\lambda_{est}$ spanned nearly the entire range of possible values (e.g., $n=32$; $\lambda_{in}=0.5$: $\lambda_{est}= 0.0\to 0.985$), revealing that estimates of $\lambda$ were not a reliable indicator of input phylogenetic signal. Importantly, the precision of $\lambda_{est}$ was not uniform across all levels of phylogenetic signal. The worst precision was observed at intermediate levels of phylogenetic signal ($\lambda_{in}\approx0.5$), while precision improved as input levels approached the extremes of $\lambda$'s range (i.e., $\lambda_{in}\rightarrow0$ & $\lambda_{in}\rightarrow1$). Thus, estimates of $\lambda$ were least reflective of the true input signal at intermediate values. Additionally, even at large levels of species richness, we found that the range of $\lambda_{est}$ still encompassed a substantial portion of possible values (e.g., $n=512$; $\lambda_{in}=0.5$: $\lambda_{est} = 0.32\to 0.68$). Likewise, the same $\lambda_{est}$ could be obtained from datasets containing vastly different input levels of phylogenetic signal (e.g., $n=512$; $\lambda_{est} = 0.5$; $\lambda_{in} = 0.25 \to 0.65$). These findings were particularly unsettling when considered in light of our literature survey. Over one quarter of the $\lambda$ estimates obtained in empirical studies (421 of 1,572) were between $\lambda=0.25$ and $\lambda=0.75$ (Figure 1). This range reflected the region that our simulations identified as being the least reliable in terms of accurately characterizing levels of phylogenetic signal, yet 30% of these empirical estimates were explicitly interpreted in terms of the strength of phylogenetic signal that they represented (i.e., weak, intermediate, strong phylogenetic signal). \hfill\break

Finally, when $\lambda$ was co-estimated with regression parameters in PGLS regression, the results of our simulations were quite similar. Here, regression parameters ($\beta$) were accurately estimated, confirming earlier findings of Revell [-@Revell2010] (see Supporting Information). However, estimates of phylogenetic signal ($\lambda$) were less precise (Figure 3), and the spread of $\lambda_{est}$ was similar to that observed when $\lambda$ was estimated for only the dependent variable, as in Figure 2. Taken together, these findings reveal that $\lambda_{est}$ does not precisely characterize observed levels of phylogenetic signal in phenotypic datasets, and that biological interpretations of the strength of phylogenetic signal based on $\lambda$ may be highly inaccurate. \hfill\break

@@

The results above demonstate that $\lambda$ is not a reliable estimate of the phylogenetic signal in phenotypic data. As such, biological interpretations of the strength of phylogenetic signal, and comparisons of the magnitude of such effects across datasets are severely compromised when based on this parameter. As an alternative, we propose that summary estimates of phylogenetic signal be converted to effect sizes for interpretation and comparison of the relative strength of phylogenetic signal in phenotypic datasets. Statistically, a standardized effect size may be found as:

@@

where $\theta_{obs}$ is the observed test statistic, $E(\theta)$ is its expected value under the null hypothesis, and $\sigma_\theta$ is its standard error [@Glass1976; @Cohen1988;@Rosenthal1994]. $Z_{\theta}$ expresses the magnitude of the effect in $\theta_{obs}$ by transforming the original test statistic to its standard normal deviate [@Glass1976; @Kelley2012]. Typically, $\theta_{obs}$ and $\sigma_\theta$ are estimated from the data, while $E(\theta)$ is obtained from the distribution of $\theta$ derived from parametric theory. However, recent advances in resampling theory [@Collyer_et_al2015a;@AdamsCollyer2016; @AdamsCollyer2019b] have shown that $E(\theta)$ and $\sigma_\theta$ may also be obtained from an empirical sampling distribution of $\theta$ obtained from permutation procedures. \hfill\break

Adams and Collyer [-@AdamsCollyer2019] suggested that the strength of phylogenetic signal could be represented as an effect size, based on the $Kappa$ statistic and its empirical sampling distribution from permutation. Here we formalize that suggestion, resulting in an effect size of:

@@

In this case, $\lambda_{obs}$ and $\hat\sigma_\lambda$ are empirically derived using maximum likelihood. Note that under the null hypothesis, no phylogenetic signal is expected [@Freckleton_et_al2002], and thus $E(\lambda)=0$ under this condition. \hfill\break

To evaluate the utility of $Z_K$ and $Z_\lambda$ we calculated both effect sizes for the simulated datasets generated above, and summarized the precision of each using its variance ($\sigma^2_{Z_K}$ and $\sigma^2_{Z_\lambda}$). Results are found in Figure 4 (additional results are found in the Supporting Information). Here two things are evident. First, estimates of $Z_K$ track the input phylogenetic signal in a more linear fashion than do estimates of $Z_\lambda$ (Figure 4A,B). Thus, actual changes in the strength of phylogenetic signal are reflected more evenly in the corresponding values of the effect size $Z_K$. Second, the precision of $Z_K$ is considerably more stable as compared with $Z_\lambda$. This may be seen by calculating the coefficients of variation for the set of precision estimates (i.e., $\sigma^2_{Z_K}$ and $\sigma^2_{Z_\lambda}$) across input levels of phylogenetic signal.  Here coefficients of variation in the precision of $Z_K$ were an order of magnitude smaller for than for $Z_\lambda$ (Figure 4C,D). This implied that estimates of the strength of phylogenetic signal were more reliable and robust when using $Z_K$ as compared with $Z_\lambda$. \hfill\break

@@

One important advantage of the approach advocated here is that the resulting effect sizes ($Z_K$) are dimensionless, as the units of measurement cancel out during the calculation of $Z$ [@SokalRohlf2012]. As a consequence, the effect sizes represent the strength of phylogenetic signal on a common and comparable scale, measured in standard deviation units, regardless of the initial units and scale of the original phenotypic variables under investigation. This means that conceivably, the strength of phylogenetic signal may be compared across datsets for phenotypic traits in different units and scale (e.g., behavioral and morphological traits), so long as they are continuous variables. Note that this is directly related to the common practice of converting data to their standard normal deviates prior to their use in multivariate analysis, if the original variables are expressed in different units or scales [see @Everitt2011; @LegendreLegendre2012; @AdamsCollyer2019].

@@

To demonstrate the utility of $\hat{Z}_{12}$ we performed an analysis of the strength of phylogenetic signal in two phenotypic datasets from species of plethodontid salamander. The data were part of a series of studies examining macroevolutionary trends in phenotypic diversification in this group [@Baken2019; @Baken2020]. Our dataset contained surface area to volume ratios (SA:V) and relative body width ($\frac{BW}{SVL}$) for 305 species (Figure 5A). For SA:V, 11 linear body measurements from 2,781 individuals were taken, from which estimates of the surface area and volume of the head, body, and tail were calculated and subsequently combined to arrive at the SA:V for each indiviual [for mathematical details see @Baken2020]. Species means were then obtained. Likewise, body size (SVL) and body width (BW) measurements were taken from 3,371 individuals, and species means of relative body width ($\frac{BW}{SVL}$) were calculated [data from @Baken2019]. A time-dated molecular phylogeny for the group [@Bonett2017] was then pruned to match the species in the dataset, resulting in a phylogeny and corresponding phenotypic dataset containing 305 species. The phylogenetic signal in each trait was then characterized using $Kappa$, which was converted to its effect size ($Z_K$) using `geomorph` 3.2.1 [@AdamsOtarola2013; @AdamsGeomorph]. Finally, the strength of phylogenetic signal was compared across traits using $\hat{Z}_{12}$ as described above (**to be incorporated in `geomorph` upon manuscript acceptance**). \hfill\break

*Results.* Both SA:V and relative body width displayed significant phylogenetic signal ($Kappa_{SA:V}=0.7608$; $P=0.001$; $Kappa_{BW/SVL}=0.2515$; $P=0.001$).  For both phenotypic traits, $K_{obs}$ differed markedly from their corresponding permutation distributions, which were found to overlap almost perfectly (Figure 5B). However, while both traits displayed significant phylogenetic signal, there was nearly a four-fold difference in the magnitude of their effect sizes, with SA:V displaying the greater phylogenetic signal (Figure 5C). Using the two-sample test statistic above, this difference was found to be highly significant ($\hat{Z}_{12}=4.13$; $P=0.000036$). Thus it may be concluded that SA:V displays significantly stronger phylogenetic signal than does relative body width, and that shared evolutionary history has strongly influenced trait covariation among taxa for SA:V. Biologically, this observation corresponds with the fact that tropical species -- which form a monophyletic group within plethodontids -- display greater variation in SA:V which covaries with disparity in their climatic niches [@Baken2020]. Thus, because of this macroevolutionary association, strong phylogenetic signal in SA:V is to be expected.
